---
title: "Tibbles in a tibble"
description: |
  Using tibble list variables in dplyr: selecting and filtering.
draft: true
base_url: https://www.psyctc.org/psyctc/Rblog/
preview: https://www.psyctc.org/psyctc/wp-content/uploads/2026/01/autocorrelation.png
categories:
  - Tidyverse
  - R programming
  - R tricks
author:
  - name: Chris Evans
    url: https://www.psyctc.org/R_blog/
    affiliation: PSYCTC.org
    affiliation_url: https://www.psyctc.org/psyctc/
    orcid_id: 0000-0002-4197-4202
date: 04-22-2026
output:
  distill::distill_article:
    toc: true
    toc_depth: 4
    hightlight_downlit: true
    self_contained: false
    code_folding: true
    includes: 
      in_header: ../../Gurgle.Rhtml
creative_commons: CC BY-SA    
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
library(tidyverse)
library(CECPfuns)
library(flextable)

### set ggplot defaults
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = .5),
             plot.subtitle = element_text(hjust = .5))
```
```{r, xaringanExtra-clipboard, echo=FALSE}
### this is just the code that creates the "copy to clipboard" function in the code blocks
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clone fa-2x\" style=\"color: #301e64\"></i>",
    success_text = "<i class=\"fa fa-check fa-2x\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times fa-2x\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

I not infrequently find myself wanting to run the same analysis, for example, computing internal reliability, for a number of measures in the same dataset and for different subsets of the data, for example by gender or other sociodemographic variables.  One way to do this that I have come to like is to use what I call "tibbles in a tibble" to contain the various selections from the one dataset.





# Summary

This may seem very convoluted but I find it's a nice way to get repeated analyses of subsets of both variables and filtered rows from a dataset. The bit I forget is how to do the filtering.


<center>Visit count
<div id="sfc27sss8g11crkt1e8bbmgtwdqaslx67b6"></div>
<script type="text/javascript" src="https://counter1.optistats.ovh/private/counter.js?c=27sss8g11crkt1e8bbmgtwdqaslx67b6&down=async" async></script>
<br><a href="https://www.freecounterstat.com">page counters</a><noscript><a href="https://www.freecounterstat.com" title="page counters"><img src="https://counter1.optistats.ovh/private/freecounterstat.php?c=27sss8g11crkt1e8bbmgtwdqaslx67b6" border="0" title="page counters" alt="page counters"></a></noscript>
</center>

## Last updated{.appendix}
```{r,echo=FALSE}
cat(paste(format(Sys.time(), "%d/%m/%Y"), "at", format(Sys.time(), "%H:%M")))
```
