---
title: "Bug in facetted ggplot with coord.fixed()"
description: |
  I think this is a bug
author:
  - name: Chris Evans
    url: https://www.psyctc.org/R_blog/
    affiliation: PSYCTC.org
    affiliation_url: https://www.psyctc.org/psyctc/
    orcid_id: 0000-0002-4197-4202
date: 2024-07-10
categories:
  - Psychometrics
  - Correlation
  - n=1
output:
  distill::distill_article:
    toc: true
    toc_depth: 4
    hightlight_downlit: true
    self_contained: false
    code_folding: true
    includes: 
      in_header: ../../Gurgle.Rhtml
creative_commons: CC BY-SA
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE, fig.width = 16, fig.height = 16)

library(tidyverse)

### set ggplot theme defaults
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = .5), 
             plot.subtitle = element_text(hjust = .5),
             text=element_text(size=24)) #change font size of all text # for knitting
             # text=element_text(size=16)) #change font size of all text # for ggsaving
             # axis.text=element_text(size=20), #change font size of axis text
             # axis.title=element_text(size=20), #change font size of axis titles
             # # plot.title=element_text(size=20), #change font size of plot title
             # legend.text=element_text(size=20), #change font size of legend text
             # legend.title=element_text(size=20)) #change font size of legend title   

### this is just the code that creates the "copy to clipboard" function in the code blocks
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clone fa-2x\" style=\"color: #301e64\"></i>",
    success_text = "<i class=\"fa fa-check fa-2x\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times fa-2x\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)

if(interactive()) {
  postDir <- "./_posts/2024-04-20-convergent-validity-issues-1"
} else {
  postDir <- "./"
}
```

# Explanation

The issue is that I get a lot of unwanted white space above and below facetted plots but not with plots without the facetting.

```{r makeData}
valN <- 2000
valCorr <- .7
matCov <- matrix(c(1, valCorr, valCorr, 1), byrow = TRUE, ncol = 2)
vecMeans <- c(0, 0)
MASS::mvrnorm(n = valN, mu = vecMeans, Sigma = matCov) %>%
  as_tibble(.name_repair = "universal") %>%
  rename(Pre = `...1`,
         Post = `...2`) %>%
  mutate(ID = row_number(),
         group = ID %% 2) -> tibDat
```

## Simple plot: no unnecessary white space.

```{r plot1}
ggplot(data = tibDat,
       aes(x = Pre, y = Post)) +
  geom_point(alpha = .1) +
  geom_smooth() +
  xlim(c(-4, 4)) +
  ylim(c(-4, 4)) +
  coord_fixed()
```

## Add facets: no unwanted white space.

```{r}
ggplot(data = tibDat,
       aes(x = Pre, y = Post)) +
  facet_grid(cols = vars(group),
             space = "free",
             ) +
  geom_point(alpha = .1) +
  geom_smooth() +
  xlim(c(-4, 4)) +
  ylim(c(-4, 4)) 
```

## Adding `coord_fixed()` causes the issue

This happens after adding facetting *and* adding `coord_fixed()`.

```{r}
ggplot(data = tibDat,
       aes(x = Pre, y = Post)) +
  facet_grid(cols = vars(group),
             space = "free",
             ) +
  geom_point(alpha = .1) +
  geom_smooth() +
  xlim(c(-4, 4)) +
  ylim(c(-4, 4)) +
  coord_fixed()
```

## Same with `theme(aspect.ratio = 1)`

There was a suggestion on the web, but about a slightly different issue about `scales = "free"` in the facetting that using `theme(aspect.ratio = 1)` was better than `coord_fixed()`.  That doesn't change things for me.

```{r}
ggplot(data = tibDat,
       aes(x = Pre, y = Post)) +
  facet_wrap(facets = vars(group)) +
  geom_point(alpha = .1) +
  geom_smooth() +
  theme(aspect.ratio = 1)
```

## Using `theme(plot.margin = ...): no good

```{r}
ggplot(data = tibDat,
       aes(x = Pre, y = Post)) +
  facet_grid(cols = vars(group),
             space = "free",
             ) +
  geom_point(alpha = .1) +
  geom_smooth() +
  xlim(c(-4, 4)) +
  ylim(c(-4, 4)) +
  coord_fixed() +
  theme(plot.margin = margin(t = 0,
                             b = 0))
```

Using `theme(plot.margin = margin(t = -70,  b = -70))` also has no effect.

```{r}
ggplot(data = tibDat,
       aes(x = Pre, y = Post)) +
  facet_grid(cols = vars(group),
             space = "free",
             ) +
  geom_point(alpha = .1) +
  geom_smooth() +
  xlim(c(-4, 4)) +
  ylim(c(-4, 4)) +
  coord_fixed() +
  theme(plot.margin = margin(t = -70,
                             b = -70))
```

Set margin units explicitly with `theme(plot.margin = unit(c(-5, 5, -5, 5), "cm"))`, still no joy on the vertical space though it can be seen that the right and left margins are imposed correctly.

```{r}
ggplot(data = tibDat,
       aes(x = Pre, y = Post)) +
  facet_grid(cols = vars(group),
             space = "free",
             ) +
  geom_point(alpha = .1) +
  geom_smooth() +
  xlim(c(-4, 4)) +
  ylim(c(-4, 4)) +
  coord_fixed() +
  theme(plot.margin = unit(c(-5, 5, -5, 5), "cm"))
```

# Conclusion

To me this is a bug.  I have a horrible feeling that I did once find it before and found some bodge that got me around it but I can't remember what that was, nor when or where!  Searching the web I'm not finding the problem or a solution.

<center>Visit count</center>
<center>
<div id="sfcg81jfzsxjdygdqh6t4t15s9fsplcpwcr"></div><script type="text/javascript" src="https://counter8.optistats.ovh/private/counter.js?c=g81jfzsxjdygdqh6t4t15s9fsplcpwcr&down=async" async></script><br><a href="https://www.freecounterstat.com">widget counter</a><noscript><a href="https://www.freecounterstat.com" title="widget counter"><img src="https://counter8.optistats.ovh/private/freecounterstat.php?c=g81jfzsxjdygdqh6t4t15s9fsplcpwcr" border="0" title="widget counter" alt="widget counter"></a></noscript>
</center>

# Dates

* *10.v.24 First created.*

## Last updated{.appendix}
```{r,echo=FALSE}
cat(paste(format(Sys.time(), "%d/%m/%Y"), "at", format(Sys.time(), "%H:%M")))
```