---
title: "Welcome to these pages"
description: |
  Welcome to these pages which I hope will be useful to people using R to analyse data.
base_url: https://www.psyctc.org/psyctc/Rblog/  
author:
  - name: Chris Evans
    url: https://www.psyctc.org/R_blog/
    affiliation: PSYCTC.org
    affiliation_url: https://www.psyctc.org/psyctc/
    orcid_id: 0000-0002-4197-4202
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating blogs with Distill at:
# https://rstudio.github.io/distill/blog.html

```
Welcome to this first attempt to create a set of pages, a sort of blog, that should provide easy access to some Rmarkdown files I'm creating that show R tricks I've found useful.  There may also be occasional pages here about things I use around my R work but the real focus will be on R.  

```{r}
library(ggplot2)
library(tidyverse)
as_tibble(list(x = 1,
               y = 1)) -> tibDat

ggplot(data = tibDat,
       aes(x = x, y = y)) +
  geom_text(label = "Welcome!",
            size = 30,
            colour = "red",
            angle = 30) +
  xlab("") +
  ylab("") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()) 
```

Do contact me if you find issues with them or have questions or suggestions.  To do that, use the [contact me](https://www.psyctc.org/psyctc/contact-me/) form on my https://www.psyctc.org/psyctc/ web site.  In most browsers I think that will open in a new page and if you close it when you have sent your message I think that and most browsers will bring you back here.





