---
title: "Wisdom of years!"
description: |
  I've learned a lot about data analysis from my errors, here's what I wish I'd known earlier!
base_url: https://www.psyctc.org/psyctc/Rblog/ 
preview: "wisdom.png"
author:
  - name: Chris Evans
    url: https://www.psyctc.org/R_blog/
    affiliation: PSYCTC.org
    affiliation_url: https://www.psyctc.org/psyctc/
    orcid_id: 0000-0002-4197-4202
categories:
  - R style
  - wise practices
date: 2023-08-06
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
```
```{r makeGraphic, echo=FALSE}
library(ggplot2)
library(tidyverse)
as_tibble(list(x = 1,
               y = 1)) -> tmpDat

# png(file = "wisdom.png", type = "cairo", width = 6000, height = 4800, res = 300)
ggplot(data = tmpDat,
       aes(x = x, y = y)) +
  geom_text(label = "Wisdom <- learning from mistakes!",
            size = 10,
            colour = "red",
            angle = 30) +
  xlab("") +
  ylab("") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) 
# dev.off()
```

This is just a little post to point to a new developing page "Wisdom1"(https://www.psyctc.org/Rblog/wisdom.html) in my little Rblog site.  It's a compilation of principles and rules to myself all of which I wish I'd learned earlier and which, I believe, save me weeks of time even though, sometimes, they can add minutes, occasionally hours and, once per project (writing DAPs & and DMPs: Data Analysis Plans and Data Management Plans) they may even take days.  Very occasionally, when trying to simulate a project, they may take even longer but those, like long DAPs, may turn into papers in their own rights.

This will accumulate and I welcome comments and suggestions [contact me](https://www.psyctc.org/psyctc/contact-me/), so I've made it a page not a post and I'm just using this to flag it up.
