---
title: "Pre/post analyses"
description: |
  Teaching myself about pre/post therapy change analyses using R.
  Probably the first of a series of posts.
base_url: https://www.psyctc.org/psyctc/Rblog/
preview: https://www.psyctc.org/psyctc/wp-content/uploads/2022/02/prepost1.png
categories:
  - change scores
  - pre/post change
author:
  - name: Chris Evans
    url: https://www.psyctc.org/R_blog/
    affiliation: PSYCTC.org
    affiliation_url: https://www.psyctc.org/psyctc/
    orcid_id: 0000-0002-4197-4202
date: 02-12-2022
output:
  distill::distill_article:
    toc: true
    toc_depth: 4
    hightlight_downlit: true
    self_contained: false
    code_folding: true
---
*Started 5.ii.22, latest update 12.ii.22, still work in progress but worth mounting to illustrate some of the issues*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(tidyverse)
library(CECPfuns)
### set ggplot defaults
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = .5),
             plot.subtitle = element_text(hjust = .5))
```
# Background

I was being very slow talking with [Emily](https://www.researchgate.net/project/Young-Persons-Clinical-Outcomes-in-Routine-Evaluation-YP-CORE-Scale-psychometric-properties-and-utility) (**Dr.** Blackshaw now) about some pre/post analyses she is doing so I realised I should take my ageing brain for some gentle walking pace exercise about pre/post change analyses!

Her dataset is fairly large and has first and last session CORE scores ([CORE-10](https://www.coresystemtrust.org.uk/instruments/core-10-information/) or [YP-CORE](https://www.coresystemtrust.org.uk/instruments/yp-core-information/), analysing each dataset separately).  She has been asked to look at the impacts on change of the numbers of sessions attended, gender and age.

Doing this has been helpful to me in thinking through the challenges of testing for effects of even a very small set of predictor variables on pre/post change scores.  I hope it may be useful to others on that level of exploring the issues.  I also hope that the code both for the exploratory/descriptive graphics, and the effect testing, will be useful to others.

# Plan

This is an emerging document and I think it will spin off some separate posts, it's also a large post and at the moment splits into three sections:

1. **Generation of the data**. This is really just here to make that process transparent and provide the code but unless you are particularly interested in this sort of thing you can ignore the code and get through this section very fast.

2. **Exploration of the data**. With real data I like to "see" the data and not just get jumped into ANOVA tables.  Here I was also doing it to get a visual sense of the effects I had created and to be sure my simulation had worked.  With real data this helps see where data doesn't fit the distributional models in the analyses (usually assumptions of Gaussian distributions and linear effects).  I have modelled in a quadratic effect, a U-shaped relationship between baseline score and age, but otherwise the data are simulated so are all pretty squeaky clean so skim this too if you want but I encourage you to look at your own data carefully with these sorts of graphics before jumping to modelling.

3. **Testing for the effects**.  This was what got me into this self-answered reassurance journey.  I was hoping that effect plots would be helpful but got into the issue of interactions so this section is really still taking shape.

# Generation of the data

This code block generates the baseline scores.  I'm using Gaussian distributions which is just one of the many unrealistic aspects of this.  However, I'm really doing this to explore different ways of displaying the effects and not aspiring to verisimilitude!

```{r createBaselineData}
### these vectors create population proportions from which to sample with sample()
vecSessions <- c(rep(2, 40), # I have made these up, I have no idea how realistic they are
                 rep(3, 30),
                 rep(4, 20),
                 rep(3, 15),
                 rep(4, 10),
                 rep(5, 8),
                 rep(6, 7),
                 rep(7, 5),
                 rep(8, 3),
                 rep(9, 2),
                 rep(10, 1))
vecAge <- c(rep(15, 20), # ditto
            rep(16, 25),
            rep(17, 25),
            rep(18, 30),
            rep(19, 35),
            rep(20, 30),
            rep(21, 30),
            rep(22, 33),
            rep(23, 29),
            rep(24, 20),
            rep(25, 18))
vecGender <- c(rep("F", 63), # ditto
               rep("M", 30),
               rep("Other", 7))

nGenders <- length(vecGender)
nAges <- length(15:25) # lazy but does make tweaking the model later easier!
nSessLengths <- length(2:20) # ditto
nCells <- nGenders * nAges * nSessLengths

avCellSize <- 20 # trying to make things big enough
populnSize <- nCells * avCellSize

### build scores from a Gaussian base variable
latentMean <- 0
latentSD <- 1

### add baseline differences
### gender has female as reference vale
effBaseFemaleMean <- 0
effBaseFemaleSD <- 1
effBaseMaleMean <- -.25
effBaseMaleSD <- 1.8
effBaseOtherMean <- .35
effBaseOtherSD <- 2
### model age as a quadratic
minAge <- min(vecAge)
midAge <- mean(vecAge)
effBaseAgeMeanMult <- .03
effBaseAgeSD <- 1

### now create model for change effects
### start with noise to add to baseline score
changeFuzzMean <- .1
changeFuzzSD <- .2
### now gender effects on change
effChangeFemaleMean <- -.8
effChangeFemaleSD <- 1
effChangeMaleMean <- effChangeFemaleMean + .2 # smaller improvement for men
effChangeMaleSD <- 1.5 # more variance in male change
effChangeOtherMean <- effChangeFemaleMean - .3 # better for "other"
effChangeOtherSD <- 1

### model age as a quadratic again
effChangeAgeMeanMult <- .1
effChangeAgeMeanSD <- 1

### model effect of number of sessions as linear
minSessions = min(vecSessions)
effChangeSessionsMult <- -.15
effChangeSessionsSD  <- 1

### build the sample
set.seed(12345) # reproducible sample
### get the basics
as_tibble(list(ID = 1:populnSize,
               gender = sample(vecGender, populnSize, replace = TRUE),
               age = sample(vecAge, populnSize, replace = TRUE),
               nSessions = sample(vecSessions, populnSize, replace = TRUE),
               ### now build baseline scores
               baseLatent = rnorm(populnSize, mean = latentMean, sd = latentSD))) -> tibSimulnVars

### now use those to build baseline data
tibSimulnVars %>%
  ### create effect of gender
  mutate(gendEffect = case_when(gender == "F" ~ rnorm(populnSize, mean = effBaseFemaleMean, sd = effBaseFemaleSD),
                                gender == "M" ~ rnorm(populnSize, mean = effBaseMaleMean, sd = effBaseMaleSD),
                                gender == "Other" ~ rnorm(populnSize, mean = effBaseOtherMean, sd = effBaseOtherSD)),
         ### this is just creating factors, useful when plotting
         Age = factor(age),
         facSessions = factor(nSessions)) %>%
  rowwise() %>%
  ### create effect of age
  ### I am, a bit unrealistically, assuming that number of sessions doesn't affect baseline score (nor v.v.)
  mutate(ageEffect = effBaseAgeMeanMult * rnorm(1, 
                                                mean = (age - midAge)^2, # centred quadratic
                                                sd = effBaseAgeSD),
         first = baseLatent + gendEffect + ageEffect) %>%
  ungroup() -> tibBaselineScores
```

Now create the change scores to create the final scores.  First time around I went straight to create the final scores, easier to follow this way.

```{r generateLastScores}
tibBaselineScores %>%
  ### create basic change scores with noise to add to the baseline scores
  mutate(change =rnorm(populnSize, 
                       mean = changeFuzzMean,
                       sd = changeFuzzSD)) %>%
  ### now add effect of gender on change
  mutate(gendChangeEffect = case_when(gender == "F" ~ rnorm(populnSize, 
                                                            mean = effChangeFemaleMean, 
                                                            sd = effChangeFemaleSD),
                                      gender == "M" ~ rnorm(populnSize, 
                                                            mean = effChangeMaleMean, 
                                                            sd = effChangeMaleSD),
                                      gender == "Other" ~ rnorm(populnSize, 
                                                                mean = effChangeOtherMean, 
                                                                sd = effChangeOtherSD))) %>%
  rowwise() %>%
  ### add effect of age
  mutate(ageChangeEffect = effChangeAgeMeanMult * rnorm(1, 
                                                        mean = (age - midAge)^2, 
                                                        sd = effChangeAgeMeanSD),
         ### add effect of number of sessions
         sessionChangeEffect = effChangeSessionsMult * rnorm(1,
                                                             mean = nSessions - minSessions,
                                                             sd = effChangeSessionsSD),
         change = change + gendChangeEffect + ageChangeEffect + sessionChangeEffect,
         last = first + change) %>%
  ungroup() -> tibDat
```

I've simulated a very unrealistic dataset of total size `r prettyNum(populnSize)` with a three way gender classification and age ranging from `r min(vecAge)` to `r max(vecAge)` and numbers of sessions from `r min(vecSessions)` to `r max(vecSessions)`.

# Exploration of the data

This was to check my simulation but I think it's good practice with any dataset to explore it graphically and thoroughly yourself and ideally to make some of that exploration available to others, some in a main paper or report, some in supplementary materials anyone can get to.

First check the breakdown by the predictor variables: gender, age and number of sessions.  With a real world dataset I'd check for systematic associations between these variables, e.g. is the age distribution, or the numbers of sessions attended different by gender?  Does the number of sessions attended relate to age?  However, I built in no non-random associations so I haven't done that exploration here.

## Gender

```{r description1}
tibDat %>%
  group_by(gender) %>%
  summarise(n = n()) %>%
  mutate(nText = str_c("n = ", n),
         yPos = n * .8) -> tmpTibN

ggplot(data = tibDat,
       aes(x = gender, fill = gender))+
  geom_histogram(stat = "count") +
  geom_text(data = tmpTibN,
             aes(x = gender, y = yPos, label = nText))
```

OK.

## Age

```{r description2}
ggplot(data = tibDat,
       aes(x = age, fill = gender))+
  geom_histogram(stat = "count") +
  scale_x_continuous(breaks = vecAge, labels = as.character(vecAge))
```

I haven't created any systematic association between age and gender.

## Numbers of sessions

```{r description3}
ggplot(data = tibDat,
       aes(x = nSessions, fill = gender))+
  geom_histogram(stat = "count") +
  scale_x_continuous(breaks = vecSessions, labels = as.character(vecSessions))
```

I haven't created any systematic association between number of sessions and gender, nor with age.

## Look at distributions of scores and how these relate to predictors

Checking distributions is a bit silly here as we know I've created samples from Gaussian distributions, however with real world data really marked deviations from Gaussian distributions would clarify that caution would be needed for any tests or confidence intervals when we look at effects on change.  

Start with "first", i.e. baseline score.

```{r baselineDist1}
### using a tweak to be able to fit a Gaussian density using ggplot::geom_density() while wanting 
### counts on the y axis not density
### found the answer at
###   https://stackoverflow.com/questions/27611438/density-curve-overlay-on-histogram-where-vertical-axis-is-frequency-aka-count
### it involves using a multiplier which is n * binwidth

### this next is actually based on https://stackoverflow.com/questions/6967664/ggplot2-histogram-with-normal-curve
tmpBinWidth <- .5
tibDat %>%
  summarise(n = n(),
            mean = mean(first),
            sd = sd(first)) -> tmpTibStats

ggplot(data = tibDat,
       aes(x = first)) +
  # geom_histogram(binwidth = tmpBinWidth) #+
  geom_histogram(binwidth = tmpBinWidth) + #-> p
  stat_function(fun = function(x) dnorm(x, mean = tmpTibStats$mean, sd = tmpTibStats$sd) * tmpTibStats$n * tmpBinWidth,
    color = "green", 
    linewidth = 1) +
  geom_vline(xintercept = mean(tibDat$first),
             colour = "blue") +
  xlab("Baseline scores") +
  ggtitle("Histogram of all baseline scores",
          subtitle = "Blue vertical reference line marks mean")
```

I amused myself by adding the best Gaussian distribution fit so I can 

and tests of fit but clearly no major problem there but that's all overkill here so I haven't.  Now what about the effects of predictors?

```{r baselineDistGend}
tibDat %>%
  group_by(gender) %>%
  summarise(mean = mean(first)) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, fill = gender)) +
  facet_grid(rows = vars(gender),
             scales = "free_y") +
  geom_histogram() +
  geom_vline(data = tmpTibMeans,
             aes(xintercept = mean)) +
  ggtitle("Histogram of baseline scores against gender",
          subtitle = "Vertical reference lines mark means")
```

We can see the relationship between mean baseline score and gender.  I have set the Y axis as free, i.e. can be different in each facet of the plot, as numbers in each gender category vary a lot.

```{r baselineDistAge}
tibDat %>%
  group_by(Age) %>%
  summarise(mean = mean(first)) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, fill = Age)) +
  facet_grid(rows = vars(Age),
             scales = "free_y") +
  geom_histogram() +
  geom_vline(data = tmpTibMeans,
             aes(xintercept = mean)) +
  ggtitle("Histogram of baseline scores against age",
          subtitle = "Vertical reference lines mark means")
```

Fine!  And just to pander to obsessionality here's a facetted histogram by both age and gender.

```{r baselineDistAge2}
tibDat %>%
  group_by(Age, gender) %>%
  summarise(mean = mean(first)) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, fill = Age)) +
  facet_grid(rows = vars(Age),
             cols = vars(gender),
             scales = "free") +
  geom_histogram() +
  geom_vline(data = tmpTibMeans,
             aes(xintercept = mean)) +
  ggtitle("Histogram of baseline scores against age",
          subtitle = "Vertical reference lines mark means")
```

That's just silly! (And it's very small here, would be lovely if distill created large plots that would open if the small plot is clicked.  I think that's beyond my programming skills.)

```{r baselineDistSessions}
tibDat %>%
  group_by(nSessions) %>%
  summarise(mean = mean(first)) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, fill = nSessions)) +
  facet_grid(rows = vars(nSessions),
             scales = "free") +
  geom_histogram() +
  geom_vline(data = tmpTibMeans,
             aes(xintercept = mean)) +
  ggtitle("Histogram of baseline scores against number of sessions",
          subtitle = "Vertical reference lines mark means")
```

OK.  Now same for final scores.

```{r lastDist1}
ggplot(data = tibDat,
       aes(x = last)) +
  geom_histogram() +
  geom_vline(xintercept = mean(tibDat$last),
             colour = "blue") +
  xlab("Final scores") +
  ggtitle("Histogram of all final scores",
          subtitle = "Blue vertical reference line marks mean")
```

OK.
  
```{r lastDistGend}
tibDat %>%
  group_by(gender) %>%
  summarise(mean = mean(first)) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, fill = gender)) +
  facet_grid(rows = vars(gender),
             scales = "free_y") +
  geom_histogram() +
  geom_vline(data = tmpTibMeans,
             aes(xintercept = mean)) +
  ggtitle("Histogram of final scores against gender",
          subtitle = "Vertical reference lines mark means")
```

OK again.

```{r lastDistAge}
tibDat %>%
  group_by(Age) %>%
  summarise(mean = mean(first)) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, fill = Age)) +
  facet_grid(rows = vars(Age),
             scales = "free_y") +
  geom_histogram() +
  geom_vline(data = tmpTibMeans,
             aes(xintercept = mean)) +
  ggtitle("Histogram of final scores against age",
          subtitle = "Vertical reference lines mark means")
```

And again.

```{r lastDistAge2}
tibDat %>%
  group_by(Age, gender) %>%
  summarise(mean = mean(first)) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, fill = Age)) +
  facet_grid(rows = vars(Age),
             cols = vars(gender),
             scales = "free") +
  geom_histogram() +
  geom_vline(data = tmpTibMeans,
             aes(xintercept = mean)) +
  ggtitle("Histogram of final scores against age",
          subtitle = "Vertical reference lines mark means")
```

Still silly!

```{r lastDistSessions}
tibDat %>%
  group_by(nSessions) %>%
  summarise(mean = mean(first)) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, fill = nSessions)) +
  facet_grid(rows = vars(nSessions),
             scales = "free") +
  geom_histogram() +
  geom_vline(data = tmpTibMeans,
             aes(xintercept = mean)) +
  ggtitle("Histogram of final scores against number of sessions",
          subtitle = "Vertical reference lines mark means")
```

Getting lumpy where the cell sizes are getting small of course but fine.


#### Gender

```{r checkBaselineGender1}
### get means and bootstrap CIs for baseline gender effect
set.seed(12345) # reproducible bootstrap
suppressWarnings(tibBaselineScores %>%
                   group_by(gender) %>%
                   summarise(mean = mean(first),
                             CI = list(getBootCImean(first, 
                                                     nGT10kerr = FALSE,
                                                     verbose = FALSE))) %>%
                   unnest_wider(CI) -> tmpTibMeans)


ggplot(data = tibBaselineScores,
       aes(x = gender, y = first)) +
  geom_violin(aes(fill = gender),
              scale = "count") +
  geom_hline(yintercept = mean(tibBaselineScores$first)) +
  geom_point(data = tmpTibMeans,
             aes(y = mean)) +
  geom_linerange(data = tmpTibMeans,
                 inherit.aes = FALSE,
                 aes(x = gender,
                     ymin = LCLmean, 
                     ymax = UCLmean)) +
  ylab("Baseline score") +
  xlab("Gender") +
  ggtitle("Violin plot to check baseline gender differences",
          subtitle = "points are means, tiny vertical lines are 95% bootstrap CI of means")
```

OK.  It's not very visible but there is a small baseline gender effect and the confidence intervals are so tight that they are just about invisible.

#### Age

```{r checkBaselineAge1}
### get means and bootstrap CIs for baseline age effect
set.seed(12345) # reproducible bootstrap
suppressWarnings(tibBaselineScores %>%
                   group_by(Age) %>%
                   summarise(mean = mean(first),
                             CI = list(getBootCImean(first, 
                                                     nGT10kerr = FALSE,
                                                     verbose = FALSE))) %>%
                   unnest_wider(CI) -> tmpTibMeans)


ggplot(data = tibBaselineScores,
       aes(x = Age, y = first)) +
  geom_violin(aes(fill = Age),
              scale = "count") +
  geom_hline(yintercept = mean(tibBaselineScores$first)) +
  geom_point(data = tmpTibMeans,
             aes(y = mean)) +
  geom_linerange(data = tmpTibMeans,
                 inherit.aes = FALSE,
                 aes(x = Age,
                     ymin = LCLmean, 
                     ymax = UCLmean)) +
  ylab("Baseline score") +
  xlab("Age") +
  ggtitle("Violin plot to check baseline gender differences",
          subtitle = "points are means, tiny vertical lines are 95% bootstrap CI of means")
```

Small and very unrealistic quadratic (U shaped) effect of age on baseline scores.

### Check on change scores I've created

```{r plotChange1}
tibDat %>%
  group_by(gender) %>%
  summarise(meanFirst = mean(first),
            meanLast = mean(last),
            CIfirst = list(getBootCImean(first, 
                                         nGT10kerr = FALSE,
                                         verbose = FALSE)),
            CIlast = list(getBootCImean(last, 
                                        nGT10kerr = FALSE,
                                        verbose = FALSE))) %>%
  unnest_wider(CIfirst) %>%
  ### got to rename to avoid name collision
  rename(obsmeanFirst = obsmean,
         LCLmeanFirst = LCLmean,
         UCLmeanFirst = UCLmean) %>%
  unnest_wider(CIlast) %>%
  ### renaming now is just for clarity rather than necessity
  rename(obsmeanLast = obsmean,
         LCLmeanLast = LCLmean,
         UCLmeanLast = UCLmean) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, y = last, colour = gender, fill = gender)) +
  geom_point(alpha = .1, size = .5) +
  geom_smooth(method = "lm") +
  # geom_point(data = tmpTibMeans,
  #            aes(x = meanFirst, y = meanLast),
  #            size = 3) +
  geom_linerange(data = tmpTibMeans,
                 inherit.aes = FALSE,
                 aes(x = meanFirst,
                     ymin = LCLmeanLast,
                     ymax = UCLmeanLast)) +
  geom_linerange(data = tmpTibMeans,
                 inherit.aes = FALSE,
                 aes(y = meanLast,
                     xmin = LCLmeanFirst,
                     xmax = UCLmeanFirst)) +
  xlab("First session score") +
  ylab("Final session score") +
  ggtitle("Scatterplot of last against first scores by gender",
          subtitle = "Lines are linear regression by gender with 95% confidence intervals\nCrosshairs are 95% confidence intervals of means")
```

Not a very informative plot here but it would be important with real data to plot something like this to see whether there are markedly non-linear relationships.  Here it's just about visible that I've created slight differences in slope of final session score on first session score by gender.  I've put in the means (of first and last scores) by gender which helps remind us of the horizontal shift of the baseline score gender differences seen above.  (Cross hairs in black as the ones for the men and for the women disappear if coloured by gender.)

```{r plotchange1b}
ggplot(data = tibDat,
       aes(x = first, y = last, colour = gender, fill = gender)) +
  facet_grid(rows = vars(gender)) +
  geom_point(alpha = .1, size = .5) +
  geom_smooth(method = "lm") +
  # geom_point(data = tmpTibMeans,
  #            aes(x = meanFirst, y = meanLast),
  #            size = 3) +
  geom_linerange(data = tmpTibMeans,
                 inherit.aes = FALSE,
                 aes(x = meanFirst,
                     ymin = LCLmeanLast,
                     ymax = UCLmeanLast)) +
  geom_linerange(data = tmpTibMeans,
                 inherit.aes = FALSE,
                 aes(y = meanLast,
                     xmin = LCLmeanFirst,
                     xmax = UCLmeanFirst)) +
  xlab("First session score") +
  ylab("Final session score") +
  ggtitle("Scatterplot of last against first scores by gender",
          subtitle = "Lines are linear regression by gender with 95% confidence intervals\nCrosshairs are 95% confidence intervals of means")


```

As the main objective here is to look for major problems with the relationship between x and y variables, best to complement that with the same but facetted by gender.

OK, no issues of non-linearities there (of course they're not, I didn't model them so!)


### What about change scores themselves?

Plotting final scores against baseline is vital to look for non-linearities in the relationship but we are as interested in change as final scores.  (Actually, we're interested in both and of course they're mathematically completely linearly related but the give usefully different views on this whole issue of final score and change.)

So plot change against first score now we have seen that the relationships between first and last scores are not markedly non-linear.

```{r plotChange2}
tibDat %>%
  group_by(gender) %>%
  summarise(meanFirst = mean(first),
            meanChange = mean(change),
            CIfirst = list(getBootCImean(first, 
                                         nGT10kerr = FALSE,
                                         verbose = FALSE)),
            CIchange = list(getBootCImean(change, 
                                          nGT10kerr = FALSE,
                                          verbose = FALSE))) %>%
  unnest_wider(CIfirst) %>%
  ### got to rename to avoid name collision
  rename(obsmeanFirst = obsmean,
         LCLmeanFirst = LCLmean,
         UCLmeanFirst = UCLmean) %>%
  unnest_wider(CIchange) %>%
  ### renaming now is just for clarity rather than necessity
  rename(obsmeanChange = obsmean,
         LCLmeanChange = LCLmean,
         UCLmeanChange = UCLmean) -> tmpTibMeans

ggplot(data = tibDat,
       aes(x = first, y = change, colour = gender, fill = gender)) +
  geom_point(alpha = .1, size = .5) +
  geom_smooth(method = "lm") +
  geom_point(data = tmpTibMeans,
             aes(x = meanFirst, y = meanChange),
             size = 3) +
  geom_linerange(data = tmpTibMeans,
                 inherit.aes = FALSE,
                 aes(x = meanFirst,
                     ymin = LCLmeanChange,
                     ymax = UCLmeanChange)) +
  geom_linerange(data = tmpTibMeans,
                 inherit.aes = FALSE,
                 aes(y = meanChange,
                     xmin = LCLmeanFirst,
                     xmax = UCLmeanFirst)) +
  xlab("First session score") +
  ylab("Final session score") +
  ggtitle("Scatterplot of change (last - first) against first scores by gender",
          subtitle = "Lines are linear regression by gender with 95% confidence intervals")
```

Now the mean points show clearly both the horizontal shifts of baseline score gender differences, but also that the change scores are different.  The CIs for the female subset are so tiny they disappear but it's clear that the differences are systematic for the change scores as well as for the baseline scores.  Very slight but clear linear relationship between baseline score and change, in real life datasets I'd expect more of a relationship and that'd be an important reason for doing this plot.

#### Age

```{r plotChange3}
ggplot(data = tibDat,
       aes(x = first, y = last, colour = Age, fill = Age)) +
  geom_point(alpha = .1, size = .5) +
  geom_smooth(method = "lm") +
  xlab("First session score") +
  ylab("Final session score") +
  ggtitle("Scatterplot of last against first scores by age",
          subtitle = "Lines are linear regression by age with 95% confidence intervals")
```

Strong relationships and no obvious non-linearities but not an easy plot to read.  Facetted plot better.

```{r plotChange3b}
tibDat %>%
  mutate(xmean = mean(first)) %>% # centre on x axis
  group_by(Age) %>%
  summarise(xmean = first(xmean), # to retain that constant
            last = mean(last)) -> tmpTibMeans


ggplot(data = tibDat,
       aes(x = first, y = last, colour = Age, fill = Age)) +
  facet_grid(rows = vars(Age)) +
  geom_point(alpha = .3, size = .5) +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = mean(tibDat$first)) +
  geom_point(data = tmpTibMeans,
             inherit.aes = FALSE,
             aes(x = xmean, y = last)) +
  xlab("First session score") +
  ylab("Final session score") +
  ggtitle("Scatterplot of last against first scores by age",
          subtitle = "Lines are linear regression by age with 95% confidence intervals\nBlack reference lines are overall mean final score, points are mean by age.")
```

Main thing here is that there are no obvious nonlinearities.  I have added the overall mean as a horizontal reference and the facet (age) mean as a point so we can still see that the final score mean is related to age.

Now change scores.

```{r plotChange4}
ggplot(data = tibDat,
       aes(x = first, y = change, colour = Age, fill = Age)) +
  geom_point(alpha = .1, size = .5) +
  geom_smooth(method = "lm") +
  xlab("First session score") +
  ylab("Final session score") +
  ggtitle("Scatterplot of change (last - first) against first scores by age",
          subtitle = "Lines are linear regression by age with 95% confidence intervals")
```

@@@ put facetted plot here later, when I have time! @@@

```{r plotChange5}
ggplot(data = tibDat,
       aes(x = first, y = last, colour = facSessions, fill = facSessions)) +
  geom_point(alpha = .1, size = .5) +
  geom_smooth(method = "lm") +
  xlab("First session score") +
  ylab("Final session score") +
  ggtitle("Scatterplot of last against first scores by n(sessions)",
          subtitle = "Lines are linear regression by n(sessions) with 95% confidence intervals")
```

@@@ put facetted plot here later, when I have time! @@@

```{r plotChange6}
ggplot(data = tibDat,
       aes(x = first, y = change, colour = facSessions, fill = facSessions)) +
  geom_point(alpha = .1, size = .5) +
  geom_smooth(method = "lm") +
  xlab("First session score") +
  ylab("Final session score") +
  ggtitle("Scatterplot of change (last - first) against first scores by n(sessions)",
          subtitle = "Lines are linear regression by n(sessions) with 95% confidence intervals")
```

@@@ put facetted plot here later, when I have time! @@@

```{r plotChange7}
### get means and bootstrap CIs for effect of n(sessions) on last score
set.seed(12345) # reproducible bootstrap
suppressWarnings(tibDat %>%
                   group_by(nSessions) %>%
                   summarise(mean = mean(change),
                             CI = list(getBootCImean(change, 
                                                     nGT10kerr = FALSE,
                                                     verbose = FALSE))) %>%
                   unnest_wider(CI) -> tmpTibMeans)

ggplot(data = tibDat,
       aes(x = nSessions, y = change, colour = facSessions, fill = facSessions)) +
  geom_violin(scale = "count") +
  geom_point(data = tmpTibMeans,
             inherit.aes = FALSE,
             aes(x = nSessions, y = mean)) +
  geom_linerange(data = tmpTibMeans,
             inherit.aes = FALSE,
             aes(x = nSessions, ymin = LCLmean, ymax = UCLmean),
             size = 1) +
  geom_smooth(inherit.aes = FALSE,
    aes(x = nSessions, y = change),
    method = "lm",
    colour = "black") +
  xlab("Number of sessions") +
  ylab("Score change") +
  ggtitle("Violin plot of change (last - first) against first scores by n(sessions)",
          subtitle = "Line is linear regression with 95% confidence interval\nPoints are means with vertical lines for their bootstrap 95% confidence intervals")

ggsave("prepost1.png")
```

# Testing for the effects

Start with linear regression of final score on baseline score with all predictors and interactions.  Age as factor.

```{r lmFull}
lm(last ~ first + gender + Age + nSessions + 
     ### add two way interactions
     gender * Age + gender * nSessions + gender * Age + Age * nSessions +
     ### add three way interaction
     gender * Age * nSessions, data = tibDat) -> lisLMFull

summary(lisLMFull)

# lisLMFull$coefficients %>%
#   as_tibble() %>% # that ignores the names so ...
#   mutate(effect = names(lisLMFull$coefficients)) %>% # get them!
#   select(effect, value) %>% # more sensible order
#   rename(coefficient = value )

### hm, that's done for me in broom
broom::tidy(lisLMFull) %>%
  ### identify the order of the terms, i.e. two-way interaction has order 2 etc.
  mutate(order = 1 + str_count(term, fixed(":")),
         sig = if_else(p.value < .05, 1, 0)) -> tibLMFull

valNinteractions <- sum(tibLMFull$order > 1)
```

That's not very digestible but it is, arguably, a sensible place to start.  We can ignore the intercept really but it's not zero!

More usefully, we have a very strong effect of initial score on final score, a statistically significant effect of male gender against the reference gender (female) and no statistically significant effect of gender "other" in this saturated model.  The reference category for age is the lowest, age 15 and all the other ages show a statistically significantly different final score from that for age 15 except age 25.  Finally, in the simple effects, we have a statistically significant effect of number of sessions on final score with coefficient estimate `r tibLMFull %>% filter(term == "nSessions") %>% select(estimate) %>% pull() %>% round(3)`, i.e. a drop of about that in mean final score for every one more session attended.  (Remember the final scores here distribute between `r round(min(tibDat$last), 2)` and  `r round(max(tibDat$last), 2)` with SD `r round(sd(tibDat$last), 2)` so I appear to have modelled in a pretty small effect of nSessions.

The complication is all those statistically significant interactions in this saturated model.  We have `r nrow(tibLMFull)` terms, including the intercept,  `r sum(tibLMFull$order == 1) - 1` simple effects (ignoring the intercept) and `r valNinteractions` interactions, `r sum(tibLMFull$order == 2)` two-way interactions and `r sum(tibLMFull$order == 3)` three-way interactions.  Here's the breakdown of the numbers significant.

```{r LMfull1}
tibLMFull %>%
  group_by(order) %>%
  summarise(n = n(),
            nSignif = sum(sig),
            propn = round(nSignif / n, 3)) %>%
  pander::pander(justify = "lrrr")
```

With `r valNinteractions` the probability that *none* of them would come out statistically significant at p < .05 given a true null population model would be .95^`r valNinteractions`, i.e. `r round(.95^valNinteractions, 3)`, pretty unlikely but the challenge is to know what to do about this.  If we could treat age as linear we wouldn't have all those effects for each age other than 15 and things would be much simpler, but we know I've modelled age as having a quadratic effect.

Cheat a bit and just fit the quadratic for age by centring and then squaring age.

```{r LMAgeQuadratic}
# lm(last ~ first + gender + poly(Age, 2) + nSessions + 
#      gender * poly(Age, 2) + gender * nSessions + gender * poly(Age, 2) + poly(Age, 2) * nSessions +
#      gender * poly(Age, 2) * nSessions, 
#    data = tibDat) -> lisLMAge2

centreVec <- function(x){
  x - mean(x)
}
tibDat %>%
  mutate(ageSquared = centreVec(age),
         ageSquared = ageSquared^2,
         ### recentre to get mean zero
         ageSquared = centreVec(ageSquared)) -> tibDat

lm(last ~ first + gender + ageSquared + nSessions + 
     gender * ageSquared + gender * nSessions + gender * ageSquared + ageSquared * nSessions +
     gender * ageSquared * nSessions, 
   data = tibDat) -> lisLMAge2

summary(lisLMAge2)
```

Hm, better but no cigar!

#### Start over from simplest model and build up

Baseline of regression model.

```{r lm1}
lm(last ~ first, data = tibDat) -> lisLM1
summary(lisLM1)
```

Of course, highly significant.

Start by adding nSessions.

```{r lmsessions}
lm(last ~ first + nSessions, data = tibDat) -> lisLMsessions
summary(lisLMsessions)

anova(lisLM1, lisLMsessions)
```

Marked effect, add gender.

```{r lmGender}
lm(last ~ first + nSessions + gender + 
     first * gender + nSessions * gender, data = tibDat) -> lisLMsessionsGend
summary(lisLMsessionsGend)

anova(lisLMsessions, lisLMsessionsGend)
```

Highly significant effect of session count remains but odd effects of gender and an interaction!

```{r lmAge1}
### age effect treating age as continuous
### short cut syntax for all interactions
lm(last ~ first * nSessions * gender * age,
   data = tibDat) -> lisLMAge
summary(lisLMAge)

anova(lisLM1, lisLMAge)
```

No effect of age as it's got a quadratic effect in my model!

```{r lmAge2}
### age effect treating age as continuous but adding quadratic term
### short cut syntax for all interactions again
lm(last ~ first * nSessions * gender * ageSquared,
   data = tibDat) -> lisLMAge
summary(lisLMAge2)

anova(lisLM1, lisLMAge2)

jtools::effect_plot(lisLMAge2, pred = ageSquared, interval = TRUE, rug = TRUE)
```
I've added an effect plot with "rugs" for the y and x variables.  Shows clear quadratic effect of age (looks linear because we're plotting against squared age).

Basically, this is a surprisingly real world mess!  I will stop here as I want to check my hunch that I've created these (as I say, very real world) interactions by the way I created the final scores using a multiplier rather than a simple addition.  However, this does demonstrate the complexities of disentangline effects with even a few predictors particularly when gender is treated not as binary and when age cannot be treated as a linear variable as it clearly has a quadratic effect.  
