---
title: "Adding best Gaussian fit to count histogram"
description: |
  How to add best Gaussian fit to a ggplot histogram based on counts not density
author:
  - name: Chris Evans
    url: https://www.psyctc.org/R_blog/
date: 2022-02-12
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

n <- 543 # sample size
```

This just to remind myself how to do something that is perhaps a bit devious: to fit the best fitting Gaussian to a histogram of an empirical distribution in ggplot.  This is typically easy in ggplot as long as you are happy with the y axis plotting density, i.e. the proportion of the total empirical distribution in each histogram bin.  The density must sum to 1. This is what Here's a trivial example based on simulating 


```{r tmp}
set.seed(12345) # get reproducible sample
rnorm(n) %>%
  as_tibble() -> tibDat

nBins <- 30
ggplot(data = tibDat,
       aes(x = value)) +
  geom_histogram(aes(y = ..density..),
                 bins = nBins,
                 center = mean(tibDat$value)) +
  stat_function(fun = dnorm,
                args(list(mean = mean(tibDat$value),
                          sd = sd(tibDat$value))),
                colour = "green") +
  geom_vline(xintercept = mean(tibDat$value),
             colour = "blue",
             size = 1.5) +
  ggtitle("Histogram of simulated Gaussian data with best fit Gaussian density curve",
          subtitle = str_c("n = ",
                           n, 
                           ", blue reference line marks mean"))
```

That's fine if you are happy having density on the y axis but I want counts and that seems to be tricky.

```{r bodge}
tmpMaxDensity <- .5 # assume the y axis density goes up to .5, may have to tweak this if it goes higher
tmpNYbreaks <- 10 # how many breaks you want on the y axis
### now use the pretty() function to give best points between 0 and n * tmpMaxDensity,
### i.e. a number you are assuming will be a bit higher than the actual highest count in any bin
tmpYpoints <- pretty(0:(n * tmpMaxDensity), tmpNYbreaks)
### but we are still actually plotting using density so convert those to fractions
tmpBreaks <- tmpYpoints / n
tmpLabels <- as.character(tmpYpoints) # labels must be character

### now the same code as above to get the histogram based on density
### but replace the y axis with a one based on counts
ggplot(data = tibDat,
       aes(x = value)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 center = mean(tibDat$value)) +
  stat_function(fun = dnorm,
                colour = "green") +
  ### here is the rescaling of the y axis using the values above
  scale_y_continuous(breaks = tmpBreaks,
                     labels = tmpLabels) +
  ylab("Count") +
  geom_vline(xintercept = mean(tibDat$value),
             colour = "blue",
             size = 1.5) +
  ggtitle("Histogram of simulated Gaussian data with best fit Gaussian density curve",
          subtitle = str_c("y axis has counts, n = ",
                           n, 
                           ", blue reference line marks mean"))
```

Now let's look at some markedly non-Gaussian data, here I have just squared the Gaussian sample data values. First the density plot.

```{r xSquared}
tibDat %>% 
  mutate(xSquared = value^2) -> tibDat

ggplot(data = tibDat,
       aes(x = xSquared)) +
  geom_histogram(aes(y = ..density..),
                 bins = nBins,
                 center = mean(tibDat$xSquared)) +
  stat_function(fun = dnorm,
                args(list(mean = mean(tibDat$xSquared),
                          sd = sd(tibDat$xSquared))),
                colour = "green") +
  geom_vline(xintercept = mean(tibDat$xSquared),
             colour = "blue",
             size = 1.5) +
  ggtitle("Histogram of simulated non-Gaussian data with best fit Gaussian density curve",
          subtitle = str_c("n = ",
                           n, 
                           ", blue reference line marks mean"))
```

Count plot.  We can see from above that we have to tweak the rescaling as the biggest bin has over 90% of the data

```{r xSquared2}
tmpMaxDensity <- .95 # assume the y axis density goes up to .5, may have to tweak this if it goes higher
tmpNYbreaks <- 10 # how many breaks you want on the y axis
### now use the pretty() function to give best points between 0 and n * tmpMaxDensity,
### i.e. a number you are assuming will be a bit higher than the actual highest count in any bin
tmpYpoints <- pretty(0:(n * tmpMaxDensity), tmpNYbreaks)
### but we are still actually plotting using density so convert those to fractions
tmpBreaks <- tmpYpoints / n
tmpLabels <- as.character(tmpYpoints) # labels must be character

### now the same code as above to get the histogram based on density
### but replace the y axis with a one based on counts
ggplot(data = tibDat,
       aes(x = xSquared)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 center = mean(tibDat$xSquared)) +
  stat_function(fun = dnorm,
                args(list(mean = mean(tibDat$xSquared),
                          sd = sd(tibDat$xSquared))),
                colour = "green") +
  ### here is the rescaling of the y axis using the values above
  scale_y_continuous(breaks = tmpBreaks,
                     labels = tmpLabels) +
  ylab("Count") +
  geom_vline(xintercept = mean(tibDat$xSquared),
             colour = "blue",
             size = 1.5) +
  ggtitle("Histogram of simulated non-Gaussian data with best fit Gaussian density curve",
          subtitle = str_c("y axis has counts, n = ",
                           n, 
                           ", blue reference line marks mean"))
```

And now the natural log transformed data (after adding 5 to all the values to avoid trying to take logs of negative numbers).


```{r ln}
tibDat %>% 
  mutate(lnX = log(value + 5)) -> tibDat

ggplot(data = tibDat,
       aes(x = lnX, y = ..scaled..)) #+
  stat_bin(aes(y = ..density..))
  geom_histogram(aes(y = ..scaled..),
                 bins = nBins) +
  geom_vline(xintercept = mean(tibDat$lnX),
             colour = "blue",
             size = 1.5) +
  stat_function(fun = dnorm,
                args(list(mean = mean(tibDat$lnX),
                          sd = sd(tibDat$lnX))),
                colour = "green") 

  ggtitle("Histogram of simulated non-Gaussian data with best fit Gaussian density curve",
          subtitle = str_c("n = ",
                           n, 
                           ", blue reference line marks mean"))
```

Count plot.  We can see from above that we have to tweak the rescaling as the biggest bin has over 90% of the data

```{r ln2}
tmpMaxDensity <- .95 # assume the y axis density goes up to .5, may have to tweak this if it goes higher
tmpNYbreaks <- 10 # how many breaks you want on the y axis
### now use the pretty() function to give best points between 0 and n * tmpMaxDensity,
### i.e. a number you are assuming will be a bit higher than the actual highest count in any bin
tmpYpoints <- pretty(0:(n * tmpMaxDensity), tmpNYbreaks)
### but we are still actually plotting using density so convert those to fractions
tmpBreaks <- tmpYpoints / n
tmpLabels <- as.character(tmpYpoints) # labels must be character

### now the same code as above to get the histogram based on density
### but replace the y axis with a one based on counts
ggplot(data = tibDat,
       aes(x = xSquared)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 center = mean(tibDat$value)) +
  stat_function(fun = dnorm,
                colour = "green") +
  ### here is the rescaling of the y axis using the values above
  scale_y_continuous(breaks = tmpBreaks,
                     labels = tmpLabels) +
  ylab("Count") +
  geom_vline(xintercept = mean(tibDat$value),
             colour = "blue",
             size = 1.5) +
  ggtitle("Histogram of simulated non-Gaussian data with best fit Gaussian density curve",
          subtitle = str_c("y axis has counts, n = ",
                           n, 
                           ", blue reference line marks mean"))
```