---
title: "Distribution overlays in ggplot: after_stat(), stat_function() and facets"
description: |
  There are many ways of overlaying a curve on a plot, typically a Gaussian on a histogram.  This is my lastest best way!
base_url: https://www.psyctc.org/psyctc/Rblog/
categories:
  - ggplot
  - Distributions
  - Graphics
  - Plots and graphs
  
author:
  - name: Chris Evans
    url: https://www.psyctc.org/R_blog/
    affiliation: PSYCTC.org
    affiliation_url: https://www.psyctc.org/psyctc/
    orcid_id: 0000-0002-4197-4202
date: 2025-05-04
output:
  distill::distill_article:
    toc: true
    toc_depth: 4
    hightlight_downlit: true
    self_contained: false
    code_folding: true
    includes: 
      in_header: ../../Gurgle.Rhtml
creative_commons: CC BY-SA
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
library(tidyverse)

### set ggplot defaults
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = .5),
             plot.subtitle = element_text(hjust = .5))
```
```{r, xaringanExtra-clipboard, echo=FALSE}
### this is just the code that creates the "copy to clipboard" function in the code blocks
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clone fa-2x\" style=\"color: #301e64\"></i>",
    success_text = "<i class=\"fa fa-check fa-2x\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times fa-2x\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

# Introduction

In an earlier post here ([Explore distributions with plots](https://www.psyctc.org/Rblog/posts/2023-11-18-explore-distributions-with-plots/)) I superimposed a Gaussian distribution curve on a histogram and the method used there, of creating a tibble with a Gaussian distribution using `dnorm()` and then superimposing a plot of that distribution on the histogram does work but is a bit tedious if you want to facet your plot.  This method does adapt to facetting fine using `stat_function()` and reminds me about using `after_stat()` rather than the denigrated double dot "..density.." syntax (many internet pages about using ggplot to superimpose distribution lines on plots still show the `..density..` syntax).

It started with this code which I used to create an [OMbook glossary](https://www.psyctc.org/psyctc/book/glossary/) entry: [“The” seven number summary](https://www.psyctc.org/psyctc/glossary2/the-seven-number-summary/).

```{r sevenNumber1}
### set sample size
n <- 5000
### set seed for reproducible results
set.seed(12345)
### create tibble of data
tibble(x = rnorm(n, mean = 10)) %>%
  ### mutate to cube those values
  mutate(xcubed = x^3,
         ### and transform to mean 10 and SD 1
         xcubed = 10 + (xcubed - mean(xcubed)) / sd(xcubed)) -> tmpTib

### pivot longer to get in form that's nice for facetting
tmpTib %>%
  pivot_longer(cols = everything(), names_to = "variable") -> tmpTibLong


### this is a key bit for use of stat_function(): 
### get  the statistics for each variable
tmpTibLong %>%
  group_by(variable) %>%
  summarise(mean = mean(value),
            median = median(value),
            sd = sd(value)) %>%
  ungroup() -> tmpTibStats


### now the plot!
ggplot(data = tmpTibLong,
       aes(x = value)) +
  ### facet by variable of interest
  facet_grid(rows = vars(variable)) +
  ### now use after_stat() to get density not count
  ### that's sensible given the aim to superimpose the
  ### Gaussian distribution curve fitting each variable
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50,
                 fill = "grey") +
  ### now use stat_function() to generate the curves 
  ### using the means and sds per variable 
  ### and fitting with the facetting
  stat_function(data = tmpTibStats,
                inherit.aes = FALSE,
                fun = dnorm,
                ### one thing that surprised me was that I had to use
                ### the explicit "tmpTibStats$" here
                args = list(mean = tmpTibStats$mean,
                            sd = tmpTibStats$sd),
                ### this tells stat_function() the number of points to 
                ### create
                            n = 1500,
                colour = "red") +
  ### very ordinary stuff from here on!
  geom_vline(data = tmpTibStats,
             aes(xintercept = mean),
             colour = "green") +
  geom_vline(data = tmpTibStats,
             aes(xintercept = median),
             colour = "blue") +
  ggtitle("Histograms of Gaussian and standardised cubed Gaussian distributions",
          subtitle = "Blue reference lines mark means, green lines mark medians")
```


There are a couple of slightly odd things to me in there:

* I had to use the explicit "tmpTibStats$" syntax to avoid getting an error message about supplying a non-numeric argument to a mathematical function in `stat_function()`.  I had thought I could just name the variable directly, i.e. "mean" rather than "tmpTibStats$mean" or that I could use `vars(mean)` but neither worked.

* The superimposed curve seems to have non-constant thickness, using the "n = 1500" has some impact on this. Low values unsurprisingly give bumpy lines and very high values don't remove the thickening where the values from dnorm() are changing more slowly.  On balance I don't dislike the non-constant thickness but I guess if it offends you you have to go back to generating Gaussian density distributions before running `ggplot()` and pulling the data in as I was doing in my [Explore distributions with plots](https://www.psyctc.org/Rblog/posts/2023-11-18-explore-distributions-with-plots/) post here.  Clearly you can do that and create a set of `dnorm()` distributions that will fit with the facetting.  Here's how I have done that for this particular example dataset.

```{r sevenNumber2}
### find the limits of the values in the datasets to know what range of 
### of Gaussian density values you need to generate
tmpTibLong %>%
  summarise(minX = min(value),
            maxX = max(value)) -> tmpTibXlimits
### how many points do you want
### I have chose this to match what I had in stat_function() above
nPoints <- 1500

### now generate those density values for each variable
tmpTibStats %>%
  ### for each valuable (not really necessary here but would be in general
  ### where we would be comparing different variables not ones standardised
  ### to have essentially the same mean and SD)
  group_by(variable) %>%
  mutate(xVals = list(seq(tmpTibXlimits$minX, 
                          tmpTibXlimits$maxX, 
                          length.out = nPoints))) %>%
  ungroup() %>%
  ### unnest to get in long format
  unnest_longer(xVals) %>%
  ### prune to only the variables we need
  select(variable, mean, sd, xVals) %>%
  ### generate the Gaussian density values
  ### using the observed mean and SD for each variable
  mutate(dGauss = dnorm(xVals, 
                        mean = mean, 
                        sd = sd)) -> tmpTibLineValues


### now do the plot!
ggplot(data = tmpTibLong,
       aes(x = value)) +
  ### facet by variable of interest
  facet_grid(rows = vars(variable)) +
  ### now use after_stat() to get density not count
  ### that's sensible given the aim to superimpose the
  ### Gaussian distribution curve fitting each variable
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50,
                 fill = "grey") +
  ### this replaces the stat_function() call
  ### and uses the Gaussian data generated above
  ### to get the density lines per facet
  geom_line(data = tmpTibLineValues,
            aes(x = xVals, y = dGauss),
            colour = "red") +
  ### very ordinary stuff from here on!
  geom_vline(data = tmpTibStats,
             aes(xintercept = mean),
             colour = "green") +
  geom_vline(data = tmpTibStats,
             aes(xintercept = median),
             colour = "blue") +
  ggtitle("Histograms of Gaussian and standardised cubed Gaussian distributions",
          subtitle = "Blue reference lines mark means, green lines mark medians")
```

# Summary/moral

There are many ways to do most things in R and sometimes I forget the one I used last and keep looking it up and finding outdated information on the internet so here's this wityh
two slightly cosmetically different ways of superimposing a density curve on a histogram.

# History

* *5.v.25: created.*

<center>Visit count</center>
<center>
<div id="sfc8r3zm38eaaxh1wm38fupw6fync9hfr12"></div>
<script type="text/javascript" src="https://counter4.optistats.ovh/private/counter.js?c=8r3zm38eaaxh1wm38fupw6fync9hfr12&down=async" async></script>
<br><a href="https://www.freecounterstat.com">web counter</a><noscript><a href="https://www.freecounterstat.com" title="web counter"><img src="https://counter4.optistats.ovh/private/freecounterstat.php?c=8r3zm38eaaxh1wm38fupw6fync9hfr12" border="0" title="web counter" alt="web counter"></a></noscript>
</center>

## Last updated{.appendix}
```{r,echo=FALSE}
cat(paste(format(Sys.time(), "%d/%m/%Y"), "at", format(Sys.time(), "%H:%M")))
```




